---
title: "Event affect live and economic across US between 1950 to 2011"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#SYNOPSIS
The data is collected by NOAA, the data is about the different major weather event,the aim of this analysis was to find, which event has more impact
on health and economic so that proper step can be taken.Data from NOAA was containing unit of money in varing range so we converted into common for analysis.
During the analysis we saw that although flood was measure cause of economic issue but not health one.
We are reading complete data but for analysis only subset of it was selected.


Loading Library need
```{r message=FALSE,warning=FALSE}
library(data.table)
library(tidyverse)
library(lubridate)
```

#Download the file
```{r}
if(!file.exists("repdata%2Fdata%2FStormData.csv")){
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "repdata%2Fdata%2FStormData.csv")
}
```


#reading raw data
```{r cache=TRUE}
(stormdata <- read_csv("repdata%2Fdata%2FStormData.csv"))
```


#take look at strcture of data
```{r}
glimpse(stormdata)
```
we can see that BGN_DATE , BGN_TIME ,END_DATE ,END_TIME are not allocated proper class so we are changing it as-
althought it is not necessary.

```{r}
stormdata$BGN_DATE <- mdy_hms(stormdata$BGN_DATE)
stormdata$END_DATE <- mdy_hms(stormdata$END_DATE)
```

the stormdata still contain unmodified BGN_TIME and END_TIME so they can be used as per use 

#select data needed for analysis (related to loss)
```{r}
economic <- stormdata %>% select(EVTYPE,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
dim(economic)
summary(economic$PROPDMG)
summary(economic$CROPDMG)
```

#let us see what kind value in our data
```{r}
unique(economic$PROPDMGEXP)
unique(economic$CROPDMGEXP)
```

there is lot of issue with unit mention in PROPDMGEXP and CROPDMGEXP.we cannot find total for now.
so lets solve unit issue. we can also see there is NA value 

#let us see what proportion of NA is available in your data
```{r}
mean(is.na(economic$CROPDMGEXP))
mean(is.na(economic$PROPDMGEXP))
```

now let solve the unit problem
#let us explore + - ? NA in our economic data
```{r}
economic %>% filter(CROPDMGEXP=="?")
economic %>% filter(PROPDMGEXP=="+")
economic %>% filter(PROPDMGEXP=="-")
economic %>% filter(PROPDMGEXP=="?")
economic %>% filter(PROPDMGEXP=="NA")
```
######let us take replace them (?,+,-) with 1 and other as per their value 

```{r}
economic %>% mutate(PROPDMGEXP=recode(PROPDMGEXP,"K"=1000,"M"=1000000,"B"=1000000000,"m"=1000000,"+"=1,"0"=1,"5"=100000,"6"=1000000,"?"=1,"4"=10000,"2"=100,"3"=1000,"h"=100,"7"=10000000,"H"=100,"-"=1,"1"=1,"8"=100000000,.default=1)) ->economic
```

```{r}
economic %>% mutate(CROPDMGEXP=recode(CROPDMGEXP,"M"=1000000,"k"=1000,"K"=1000,"m"=1000000,"B"=1000000000,"2"=100,.default=1)) ->economic
```

#let us grab a look at our data  values

```{r}
unique(economic$PROPDMGEXP)
unique(economic$CROPDMGEXP)
```


#let us replace all NA to 1

```{r}
economic$PROPDMGEXP %>% replace_na(0) ->economic$PROPDMGEXP
economic$CROPDMGEXP %>% replace_na(0) ->economic$CROPDMGEXP
```

#let us now compute the loss
```{r}
economic %>% mutate(TOTAL_PROPDMG = (PROPDMG*PROPDMGEXP),TOTAL_CROPDMG = (CROPDMG*CROPDMGEXP),TOTAL_LOSS=(TOTAL_PROPDMG+TOTAL_CROPDMG)) ->new_economic
```

#let see which event did how much loss
```{r}
new_economic %>% group_by(EVTYPE)%>%
summarise(LOSS_PER_EVENT=sum(TOTAL_LOSS))%>%
arrange(desc(LOSS_PER_EVENT)) -> economic_value
```

#plot this information 
```{r}
ggplot(data=economic_value)+geom_point(aes(EVTYPE,LOSS_PER_EVENT))
```

###we can see that value graph is not informative so let us look at top row for now

```{r}
ggplot(data=economic_value[1:4,])+geom_col(aes(EVTYPE,LOSS_PER_EVENT,fill=EVTYPE))+ggtitle(" top 4 types of events that have the greatest economic consequences")
```

#let us select data for analysis (related to health)
```{r}
 health <- stormdata %>% select(EVTYPE,FATALITIES,INJURIES)
```

#let us see at our data 
```{r}
summary(health$FATALITIES)
summary(health$INJURIES)
```

#let us see how many live were affected 
```{r}
health %>% mutate(total_life_affected =(FATALITIES+INJURIES)) -> healthdata
```

#let us see which event affected more lives than others
```{r}
healthdata %>% group_by(EVTYPE)%>%
summarise(total_life_affected = sum(total_life_affected))%>%
arrange(desc(total_life_affected))-> life_data
```

#let have look at graph to see which event was prominent
```{r}
ggplot(data=life_data[1:4,])+geom_col(aes(EVTYPE,total_life_affected,fill=EVTYPE))+ggtitle(" top 4 types of events that have the greatest health consequences")
```

#CONCLUSION
it seem that flood cause more economic harm and tornado causing live losses and injuries.
so during tornado proper measure can be taken for human safety while  during flood more attention must be paid to secure economic.